Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities

### 2021-09-29 paper/manuscript.Rmd
Brain dump in discussion, add citations for studies.

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: #word_document
  rticles::elsevier_article:
  includes:
   in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: sentence
---

@@

It is critical to establish whether or not functional loss is the general rule for communities.
While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001]. Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020]. And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020]. On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with large impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].

@@

Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change (citation).
While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
In order to ensure comparability across taxa in trait type and data quality we therefore we focus on Anamalia here.

@@

<!-- We expect to find a general trend of functional loss underlying similar patterns of constant species richness and increasing turnover, as communities exhibit the impacts of accelerating environmental change. -->

<!-- While trait data exists for other taxa in the BioTIME database, we focus on these groups because they have accepted species-level means, comparable traits collected across taxa, and near comprehensive species coverage. Plants in particular have a rich history of functional trait work, but are far more specoise and suffer from biases in trait coverage that could have meaningful impacts on interpretation of trends [@fitzjohn2014]. -->

<!--These taxa represent a significant portion of the world's biodiversity and we have no prior expectation that the inclusion of additional taxa would significantly alter the picture of functional change. W and as such should capture general trends -->

<!--We have no prior expectation based on ecological theory that broad scale functional shifts would be significantly different between taxa, as communities of all different kinds of organisms are shaped by the same processes (citation). -->

\
<!--if the hypothesis of global biodiversity trends is true, there is no reason to believe that it would not be evident in animals alone, and only manifest when plants are included. Examining trends in functional traits of plants is certainly of equal interest, but raises additional challenges (species-level means, comparable traits cross-taxa, bias).  Therefore we focus on Anamalia here.-->

@@

```{r echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  knitr::kable(format = "latex", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

@@

@@

If this is true, it has significant implications for how we approach work on biodiversity and ecosystem function.
Very few observed communities meet the criteria for biodviersity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
Need to measure the right traits, but also decide what kind of shifts are important for functional change.
But this is also good news!
Most communities appear to be just fine from a function standpoint following best practices.

@@

<!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species? It seems inconsistent with their findings --> Maintenance of functional composition contradicts likely scenarios of loss predicted due to rare species extinction.
Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020]. Assuming local extirpation is a precursor to global extinction, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
<!-- good news for ecosystem function! -->

>>>>

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: #word_document
  rticles::elsevier_article:
  includes:
   in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: sentence

nocite: |
  @bakker1990, @berezovikovn.n.2004, @enemar2004, @gido2019, @hogstad1993, @holmes1986, @jahncke2006, @krivenkov.g.1991, @melnikovy.i.2000, @monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013, @monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014, @monitori2014, @pardieck2020, @preston1960, @svensson2006, @svensson2010, @thorn2016changes, @thorn2016response, @usfs, @vermontcenterforecostudies2015, @vickery1984, @waide2017, @williamson1983, @woehlerericj., @zakharovv.d.1998, @meditss2011, @animald, @baltics, @piropno, @rossaferes1997, @scott, @silva, @ernest2009, @friggens2008, @jalilova.b.2014, @kartzinel2014, @kaufman2019, @krefting1974, @lightfoot, @malyshevy.s.2011, @meyer2008, @nedosekinv.y.2015, @prins1990, @rocha2017, @sheftelb.i., @stapp2014, @monitori2014, @bahamas, @popacet
---

@@

It is critical to establish whether or not functional loss is the general rule for communities.
While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001].
Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].

@@

Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
In order to ensure comparability across taxa in trait type and data quality we therefore we focus on Anamalia here.

@@

@@

```{r echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  kableExtra::kbl(format = "markdown", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") #%>%
  #kable_classic() #%>% 
  #kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

@@

All code for reproducing analyses are archived at (zenodo link).
Data products are archived at (zenodo link).
Raw

@@

Local maintenance of functional structure has important implications for biodiversity and ecosystem function (BEF).
BEF work links ecosystem function to changes in species and functional diversity, however very few observed communities experienced changes along either of those characteristics.
Do we then not need to worry about loss of function in most bird, mammal and amphibian communities?
Always the option that we didn't measure traits related to function, though in this case I think we did.
We did not however measure fucntional redundancy, which could be important for maintaining multifunctionality [@bagousse-pinguet2019] .
Also may not be sensitive enough to detect significant decreases in abundance for the most dominant species, which may also be important

-   Traits are not grounded in theory, might be measuring the wrong ones

-

Even if there were unobserved communities experiencing huge shifts,

meet the criteria for biodiversity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
Need to measure the right traits, but also decide what kind of shifts are important for functional change.
But this is also good news!
Most communities appear to be just fine from a function standpoint following best practices.

@@

<!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species? It seems inconsistent with their findings -->

Maintenance of functional composition contradicts likely scenarios of loss predicted due to rare species extinction.
Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
Assuming local extirpation is a precursor to global extinction, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
<!-- good news for ecosystem function! -->
